Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and attentional function: A study based on the parametric g-formula

### 2024-03-12 output/paper/manuscript.qmd
Reviewing.

format:
  docx:
    # reference-doc: reference_document.docx
    page-width: 8.3
    citeproc: true
    mermaid:
      theme: neutral
  pdf:
    documentclass: article
    geometry:
      - top=30mm
      - left=20mm
      - heightrounded
    pdf-engine: lualatex
    number-sections: true
    toc: false
    keep-tex: false
    include-in-header:
      text: |
        \usepackage{lineno}
        \usepackage{rotating}
        \usepackage{float}
        \usepackage{typearea}
    include-before-body:
      text: |
        \linenumbers

@@

Evidence suggests that {{< acr edcs >}} may perturb the {{< acr hpa >}} axis, which has a major role in brain development. We aimed to evaluate the effects of childhood exposure to organophosphate pesticides, phenols, and phthalate metabolites, on urinary glucocorticosteroids and inattention in children using data from the {{< acr helix >}} cohort. We used the parametric g-formula to estimate effects between {{< acr edcs >}}, glucocorticosteroids, and {{< acr hitrtse >}}, a measure of inattention from the {{< acr ant >}}, and tested for possible effect modification by sex. We observed a positive {{< acr mc >}} for exposure increases from the 10th to the 90th percentile for methyl-paraben (`r format_res(meffects1, "mepa", "HRT-SE")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "HRT-SE")`), oh-MiNP (`r format_res(meffects1, "ohminp", "HRT-SE")`), and MEHP (`r format_res(meffects1, "mehp", "HRT-SE")`), on {{< acr hitrtse >}}. Several {{< acr edcs >}} were also associated with a positive {{< acr mc >}} for cortisone, cortisol, and corticosterone production. Increased levels of the glucocorticosteroids had no effect on {{< acr hitrtse >}}, although we found a possible effect modification by sex. Our results suggest that multiple {{< acr edcs >}} might interfere with inattention in children and with the homeostasis of the {{< acr hpa >}} axis.

@@

For each research question, defined by a specific type of exposure and outcome, the minimal set of covariates for inclusion in the analyses was selected on the basis of a {{< acr dag >}} built with `DAGitty` [@TextorvanderZanderGilthorpe:2016] and `ggdag` [@Barrett:2023]. The sets of covariates were selected to estimate the total effect of the exposure on the outcome. For effect estimation of the {{< acr edcs >}} on glucocorticosteroids and of glucocorticosteroids on {{< acr hitrtse >}}, these sets were also sufficient to estimate direct effects. Sample-specific creatinine values were used to adjust for possible dilution effects. Further, each minimal adjustment set was *augmented* with precision covariates, defined as the set of parents variable of the outcome that are not parents of the exposure. The adjustment sets are provided in the Supplementary Material as text files compatible with DAGitty. Codebooks for the used covariates, by research question, are provided in Supplementary Tables [-@supptbl-codebook-1; -@supptbl-codebook-2; -@supptbl-codebook-3].

@@

Concentrations of the non-persistent {{< acr edcs >}} were classified as quantifiable, below the {{< acr lod >}}, possible interference or out of range, and not analysed. Concentrations below the {{< acr lod >}} were singly imputed using a quantile regression approach for the imputation of left-censored missing data, as implemented in the `impute.QRILC` function from the `imputeLCMD` R package [@lazar2015imputelcmd]. Information about the lower limits of detection can be found in @HaugSakhiCequier:2018. Chemicals with more than 70% of observations below the {{< acr lod >}} were not considered in the present study. Remaining missing values were imputed similarly using kNN. Values of the chemicals were expressed in $\mu$grams per litre ($\mu$g/L).

@@

Strengths of the present study include the use of pooled urine samples for chemical assessment to obtain more representative long-term exposures, since it is known that these specific {{< acr edcs >}} have very short half-lives [@PerrierGiorgis-AllemandSlama:2016; @CasasBasaganaSakhi:2018]. We decided to model both the *treatment* mechanisms, for the estimation of balancing weights, and the outcomes, with traditional covariates adjustment, to try to obtain *doubly robust* effect estimates. Finally, we decided not to interpret our results by focusing on the estimated coefficients of the regression models, but by making use of the g-computation procedure and estimate {{< acr mcs >}}.

>>>>

format:
  html: default
  docx:
    # reference-doc: reference_document.docx
    page-width: 8.3
    citeproc: true
    mermaid:
      theme: neutral
  pdf:
    documentclass: article
    geometry:
      - top=30mm
      - left=20mm
      - heightrounded
    pdf-engine: lualatex
    number-sections: true
    toc: false
    keep-tex: false
    include-in-header:
      text: |
        \usepackage{lineno}
        \usepackage{rotating}
        \usepackage{float}
        \usepackage{typearea}
    include-before-body:
      text: |
        \linenumbers

@@

Evidence suggests that {{< acr edcs >}} may perturb the {{< acr hpa >}} axis, which has a major role in brain development. We aimed to evaluate the effects of childhood exposure to organophosphate pesticides, phenols, and phthalate metabolites, on urinary glucocorticosteroids and inattention in children using data from the {{< acr helix >}} cohort. We used the parametric g-formula to estimate effects between {{< acr edcs >}}, glucocorticosteroids, and {{< acr hitrtse >}}, a measure of inattention from the {{< acr ant >}}, and tested for possible effect modification by sex. We observed a positive {{< acr mc >}} for exposure increases from the 10th to the 90th percentile for methyl-paraben (`r format_res(meffects1, "mepa", "HRT-SE")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "HRT-SE")`), oh-MiNP (`r format_res(meffects1, "ohminp", "HRT-SE")`), and MEHP (`r format_res(meffects1, "mehp", "HRT-SE")`), on {{< acr hitrtse >}}, indicating lower attention. Several {{< acr edcs >}} were also associated with a positive {{< acr mc >}} for cortisone, cortisol, and corticosterone production. Increased levels of the glucocorticosteroids had no effect on {{< acr hitrtse >}}, although we found a possible effect modification by sex. Our results suggest that multiple {{< acr edcs >}} might interfere with inattention in children and with the homeostasis of the {{< acr hpa >}} axis.

@@

For each research question, defined by a specific type of exposure and outcome, the minimal set of covariates for inclusion in the analyses was selected on the basis of a {{< acr dag >}} built with `DAGitty` [@TextorvanderZanderGilthorpe:2016] and `ggdag` [@Barrett:2023]. The sets of covariates were selected to estimate the total effect of the exposure on the outcome. For effect estimation of the {{< acr edcs >}} on glucocorticosteroids and of glucocorticosteroids on {{< acr hitrtse >}}, these sets were also sufficient to estimate direct effects. Sample-specific creatinine values were used to adjust for possible dilution effects. Further, each minimal adjustment set was *augmented* with precision covariates, defined as the set of parents variable of the outcome that are not parents of the exposure. Common confounders were cohort, ethnicity, sex, age, height, weight, and head circumference of the child, consumption of fish, fruit, vegetables, organic food, anf fast food, maternal tobacco consumption, family financial situation and affluence scale (FAS). Models for estimating the effects of {{< acr edcs >}} on {{< acr hitrtse >}} were further adjusted for child breastfeeding, prenatal maternal active and passive smoking, urine creatinine, child mood and rest before assessment, child neuropsychological diagnosis, marital status, season, and fasting time before assessment. Models for estimating the effects of {{< acr edcs >}} on glucocorticosteroids were further adjusted for urine creatinine, season, and fasting time before assessment. Models for estimating the effects of glucocorticosteroids on {{< acr hitrtse >}} were further adjusted for child breastfeeding, prenatal maternal active and passive smoking, marital status, {{< acr edcs >}}, urine creatinine, child mood and rest before assessment, and child neuropsychological diagnosis. The adjustment sets are provided in the Supplementary Material as text files compatible with DAGitty. Codebooks for the used covariates, by research question, are provided in Supplementary Tables [-@supptbl-codebook-1; -@supptbl-codebook-2; -@supptbl-codebook-3].

@@

Concentrations of the non-persistent {{< acr edcs >}} were classified as quantifiable, below the {{< acr lod >}}, possible interference or out of range, and not analysed. Concentrations below the {{< acr lod >}} were singly imputed using a quantile regression approach for the imputation of left-censored missing data, as implemented in the `impute.QRILC` function from the `imputeLCMD` R package [@lazar2015imputelcmd]. Information about the lower limits of detection can be found in @HaugSakhiCequier:2018. Chemicals with more than 70% of observations below the {{< acr lod >}} were excluded from the present study. Remaining missing values were imputed similarly using kNN. Values of the chemicals were expressed in $\mu$grams per litre ($\mu$g/L).

@@

Strengths of the present study include the use of pooled urine samples for chemical assessment to obtain more representative long-term exposures, since it is known that these specific {{< acr edcs >}} have very short half-lives [@PerrierGiorgis-AllemandSlama:2016; @CasasBasaganaSakhi:2018]. We decided to model both the *treatment* mechanisms, for the estimation of balancing weights, and the outcomes, with traditional covariates adjustment, to try to obtain *doubly robust* effect estimates. Finally, we decided not to interpret our results by focusing on the estimated coefficients of possibly misspecified regression models, but by making use of the g-computation procedure and estimate {{< acr mcs >}}.
